---
title: "A Land Use Solution in Search of a Transport Problem"
author: "Jonathan Wright @citywright"
date: "30 July 2016"
output: html_document
---

```{r Code, echo=FALSE, warning=FALSE, message=FALSE}
# ====================================================================================
# Australian Airline Passenger Time Series
# Analysis by Jon Wright @citywright
# Data from Bureau of Infrastructure, Transport and Regional Economics (BITRE)
# https://bitre.gov.au/publications/ongoing/domestic_airline_activity-time_series.aspx
# Accessed on 27 July 2016
# ====================================================================================

# This script relies on the following packages already being installed:

library(ggplot2)
library(lubridate)
library(reshape2)
library(dplyr)

# ------------------------------------------------------------------------------------
# NOTE: Due to OSx Yosemite Java error in the package "xlsx", I have downloaded the 
# data manually and converted the file manually to .csv format.  Note that only the 
# tab called "Top Routes" in each spreadsheet is used in this analysis. This script 
# assumes each .csv file is already in the local working directory. (When I have a
# little more time on my hands, I'll fix up the Java error and rewrite the download
# script so that this pulls directly from the web.)
# ------------------------------------------------------------------------------------



# -- Import -- #

# Download the following files and save tab "Top Routes" as .csv file in working directory.
# "https://bitre.gov.au/publications/ongoing/files/domestic_airline_activity_Toproutes_Jan84-Jun94.xls"
# "https://bitre.gov.au/publications/ongoing/files/domestic_airline_activity_TopRoutesJuly1994June2004.xls"
# "https://bitre.gov.au/publications/ongoing/files/Domestic_aviation_activity_TopRoutesJuly2004May2016.xls"

air_1984_1994 <- read.csv(file="domestic_airline_activity_Toproutes_Jan84-Jun94.csv",
                          skip =  6, stringsAsFactors = FALSE)
air_1994_2004 <- read.csv(file="domestic_airline_activity_TopRoutesJuly1994June2004.csv",
                          skip =  7, stringsAsFactors = FALSE)
air_2004_2016 <- read.csv(file="Domestic_aviation_activity_TopRoutesJuly2004May2016.csv",
                          skip = 11, stringsAsFactors = FALSE)



# -- Clean Up -- #

# Clear out extra header rows and append the three datasets, taking only the monthly data
air01 <- air_1984_1994[-c(1:3),]
air02 <- air_1994_2004[-1,c(1:10)]
air03 <- air_2004_2016[-1,c(1:10)]
airfull <- rbind(air01,air02,air03)

# Clean up colunm names
colnames(airfull)[1:2]   <- c("City_1","City_2")
colnames(airfull)[5:8]   <- c("RevPax","AircraftTrips","LoadFactorPCT","DistanceKM")

# Convert numbers stored as strings to integers
for (j in seq(5,10)) {
    airfull[,j] <- as.numeric(gsub(",","",airfull[,j]))
}

# Convert Origin, Destination fields to factors
for (j in c(1,2)) {
    airfull[,j] <- as.factor(airfull[,j])
}


# -- Process -- #

# Convert Year and Month into single date field (for x-axis of plots)
airfull$LongDate <- ymd(with(airfull,paste(Year,Month,"1",sep="/")))

# Create CityPair field as hyphenated pair of City_1 and City_2
airfull$CityPair <- with(airfull, paste(City_1,City_2,sep="-"))

# Create RevPax field in thousands, for plotting purposes
airfull$RevPaxK <- airfull$RevPax/1000

# Create LoadFactor field in actual percentage
airfull$LoadFactor <- airfull$LoadFactorPCT/100

```

High-speed rail has captured Australia's imagination for decades, generating studies and attracting support 
[across](http://greens.org.au/high-speed-rail) the [political](http://www.100positivepolicies.org.au/high_speed_rail) [spectrum](http://www.theaustralian.com.au/national-affairs/malcolm-turnbulls-ticket-to-deliver-fast-train/news-story/feae238d4c3a6253982e28ff4e156c94).  Sydney to Melbourne is already the [fifth-most traveled](http://www.economist.com/blogs/graphicdetail/2012/05/daily-chart-8) commercial air route in 
the world.  With all the hassles of flying, the thought of being whisked comfortably from Spencer Street to 
Circular Quay in under two hours is indeed alluring.  But sky-high construction costs and long profitability 
timelines have kept the dream out of reach.  According to the [latest government report](https://infrastructure.gov.au/rail/trains/high_speed/), high-speed rail from 
Brisbane to Melbourne could cost $114 billion and take 40 years to cover costs.

Enter Consolidated Land and Rail Australia (CLARA), a private group with [a proposal](http://www.clara.com.au/the-clara-plan.html) to deliver the first stage 
high-speed rail from Sydney to Melbourne without public funding.  All this while promising to relieve 
development pressures and unclog congestion in existing cities by “rebalancing settlement” to eight new regional
cities along the rail line.  The proposal relies on 'value capture,' an [established](http://infrastructureaustralia.gov.au/policy-publications/publications/files/IA_J16-2330_Fact_Sheet_Funding_v1.2.pdf) and [highly-touted](http://www.afr.com/opinion/value-capture-is-infrastructure-magic-bullet-20151021-gkeqj9) model for
funding public infrastructure by capitalizing on increased land values through tax or contributions.  

But the CLARA proposal bets big on three tenuous assumptions that should give the government pause: the ease 
of creating new cities, the benefits of urban decentralization, and the long-term viability of high-speed rail.  

```{r Plot, echo=FALSE, warning=FALSE}

pax <- ggplot(airfull, aes(x=LongDate, y=RevPaxK, color=CityPair)) + geom_line(alpha=0.5) +
        labs(title = "Australian Monthly Domestic Airline Passengers by City Pair, 1984-2016",
             caption = "Created in Rstudio using data from https://bitre.gov.au/publications/ongoing/domestic_airline_activity-time_series.aspx") + 
        xlab("Date") + ylab("Passengers in Thousands") + 
        theme(plot.title = element_text(size = 12, face = "bold"),
              axis.title = element_text(size = 10),
              plot.caption = element_text(size = 8),
              legend.position = "none") +
        geom_text(aes(x=airfull$LongDate[20380], y=750, label="SYD-MEL"), color="black")

plot(pax)
```

Value capture works best when supporting [sensible infrastructure](http://www.theage.com.au/victoria/highrises-to-soar-over-suburban-stations-to-help-fund-level-crossing-removals-20160725-gqd3gi.html) near existing land use assets, [like Sydney's 
Central station](https://infrastructure.gov.au/rail/trains/high_speed/files/HSR_Phase_2_Chapter_7.pdf.  If it also requires compulsory acquisition, the government must take extra care not to 
conflate private profits with public interest.  CLARA would create and capture value from new cities along the
line, the cost and complexity of which cannot be underestimated.  New cities need more than housing and a 
station.  They need roads and pipes; infrastructure for energy, waste, and communications; parks, reserves, 
schools, hospitals, and businesses.  And cities needs people to run all of these; not everyone will commune by 
train.  

So cities must also be communities – vibrant and attractive places in their own right.  Places are successful 
when they impart a sense of the timeless.  The irony of a futuristic, hyper-connected city in a paddock is that 
it would be wholly disconnected from any sense of history.  The CLARA proposal risks putting too much faith in 
our ability to create sustainable places from scratch.  Great thinkers from [Jane Jacobs](https://www.theguardian.com/cities/2016/may/04/jane-jacobs-100th-birthday-saskia-sassen) to [Jan Gehl](http://www.fastcodesign.com/3061586/slicker-city/5-rules-for-designing-great-cities-from-denmarks-star-urbanist) have taught 
that the soul of great places emerge from the unplanned and unexpected.   

If these cities fail to take off, how will later stages of the rail be funded?  What would be the public cost 
of propping these cities up, or the consequences of winding them down?  

CLARA touts high-speed rail as “the best technology available for the foreseeable future.”  But the transport 
sector is undergoing a dramatic upheaval from a [convergence of disruptive technologies](http://www.vox.com/a/new-economy-future/transportation) and the rise of the 
sharing economy.  Advances in [autonomous vehicles](https://www.wired.com/2016/01/the-metastructure-transportation/) could dramatically increase the safety and capacity of 
existing motorways, rebalancing the location equation for housing.  Advances in virtual and [augmented reality](http://www.wired.com/2016/04/magic-leap-vr/) 
could [reduce the need for travel](http://www.wired.com/2015/09/my-life-as-a-robot-double-robotics-telecommuting-longread/) altogether.  New technologies that sound [outlandish](http://www.theverge.com/2016/7/5/12099612/hyperloop-one-helsinki-stockholm-30-minutes-study) now could be [closer than we think](http://waitbutwhy.com/2015/01/artificial-intelligence-revolution-1.html).   

If the rail line fails to achieve the necessary ridership to cover costs, will the public be willing and able 
to subsidize its operations to keep it running?  

Even if we could create livable cities, it's worth asking whether we should.  Seventy years of [outward sprawl](http://www.citylab.com/housing/2015/02/why-are-developers-still-building-sprawl/385922/) 
has left us [struggling](http://www.strongtowns.org/journal/2011/6/13/the-growth-ponzi-scheme-part-1.html) to pay maintenance and mortgages.  Decades of [designing cities for cars and traffic](http://www.theatlantic.com/business/archive/2016/04/absurd-primacy-of-the-car-in-american-life/476346/), to 
paraphrase the urbanist [Fred Kent](http://www.pps.org/blog/transportationasplace/), have embedded social challenges into the fabric of our cities: from traffic 
congestion to social disconnection.  The CLARA proposal mistakenly assumes that the answer to our growing cities 
is to continue spreading further out, rather than seeking infrastructure efficiency through consolidation.

We cannot afford to [disregard the risks of success](http://mobile.nytimes.com/2016/06/29/world/europe/boris-johnson-brexit-leaders-eu.html).  If the CLARA proposal were built, what would the 
opportunity costs be for our existing cities?  Which infrastructure projects and public policies would we 
forgo by shifting money, energy, and expertise towards a high-speed rail line and eight new cities?  

It's easy to find risks in big ideas, and the CLARA proposal deserves credit for its ambition.  This shouldn't
stop governments from thinking big, for the wicked problems facing our cities – climate change, housing 
affordability, aging population, and a changing economy – will require strong leadership and bold solutions.  
But nor should the enormity of our collective challenges inspire haste.  

The government should continue to investigate transformative projects like high-speed rail, and value-capture 
should play a role.  But it should be used to support necessary infrastructure rather than land 
speculation.  

Instead, we should look to invest in solutions that bring people together rather than spread them apart, and 
plan for innovation rather than betting big on one technology.  Instead of reinventing our cities, let's 
[keep building](http://www.theatlantic.com/magazine/archive/2016/03/how-america-is-putting-itself-back-together/426882/) on what already makes them great.  